A Theory for Database Transactions with no Database

### 2019-03-04 2018 Transactions/articleTransactions.tex
introducing transactions

	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
	A {\em historic structure} is a tuple $\la \Events, \id{prec}\ra$ in which
\begin{eqnarray}
	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
	So for every event $e\in\Events$ we can say that it has happened (at some time in the past).
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\Events$.
	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
	meaning that $\id{prec}^*$ is reflexive, transitive, and antisymmetric.
	For this reason equation~\ref{req:precPlus asy} requires the transitive closure $\id{prec}^+$ to be antisymmetric.

@@

	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
	Note that any nonempty linear historic structure has one current event.
	Other historic structures may have more.

@@

	If a DIS is initialized with an empty historic structure
	and it grows by means of $\id{happens}$, i.e. by one event at a time,
	then at any moment in time it consists of a finite number of events.
	Therefore, we can safely assume that set $\Events$ is finite in every practical situation.

	In a linear historic structure,
	$P$ is alway a singleton set.
	In that case, every event except the first one has precisely one preceding event.
	In general, however, events may have multiple preceding events and the predecessor set is not necessarily a singleton set.
\begin{lemma}
\label{lemma:happens preserves historic structure}
	If $h=\la \Events, \id{prec}\ra$ is a historic structure and $P\subseteq\Events$ and $e\notin\Events$
	then $\id{happens}_{P,e}(h)$ is a historic structure.
\end{lemma}
	Lemma~\ref{lemma:happens preserves historic structure} says that $\id{happens}_{P,e}$ preserves the historic structure.
	This lemma has been proven in proof assistant Isabelle/HOL.
	To understand why $P\subseteq\Events$ is necessary,
	let $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
	However, this violates requirement~\ref{req:precInE}.
	So it is necessary that every $e_1\in P$ is an element of $\Events$.
	Similarly, $e\notin\Events$ is a necessary condition,
	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).

@@

	Having defined historic structures and the way they grow, let us proceed to determine requirements for states.
	In the sequel, we will use variables $h$ and $h'$ to denote history structures.

@@

	Let $\mathbb S$ represent the set of all conceivable states,
	with one special state $\mathcal I$ that represents the initial state.
	A {\em stateful historic structure} (SHS) is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
	$\la \Events, \id{prec}\ra$ is a historic structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
	that maps every event in $\Events$ to a state.
	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
	Or shorter: $s$ is the state after event $e$.
	Alternatively, we may write $(s,e)\in\id{state}$ because a function is a univalent and total relation
	(and therefore a set of pairs).
	A {\em current state} is the state after a current event
	and a {\em past state} is the state after a past event.

	In a stateful historic structure, the occurrence of an event $e$ requires that $\id{state}(e)$ exists.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another:
\begin{equation}
\begin{array}{l}
	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
\end{array}	
\label{eqn:def occurs}
\end{equation}
	In fact, this definition expands the definition of historic structures (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves historic structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful historic structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.

	Let $\mathbb D$ represent the set of all conceivable state changes.
	Let $s_2\ominus s_1$ represent the change from state $s_1$ to state $s_2$.
	Let $\delta(e_1,e_2)$ represent the change $\id{state}(e_2)\ominus\id{state}(e_1)$.

	Each time a new event happens, a change of state takes place.
	This change is described in the following requirements:
\begin{eqnarray}
	&&(\exists e_2:(e_1,e_2)\in\id{prec})\ \vee\ \id{state}(e_1)=\mathcal I\label{eqn:totState}\\
	&&(e_1,e_2)\in\id{prec}\ \Rightarrow\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\label{eqn:nextState}
\end{eqnarray}
	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
	Requirement~\ref{eqn:nextState} says that the state of non-initial events can be computed using $\oplus$.
	Together, requirements~\ref{eqn:totState} and~\ref{eqn:nextState} define the function $\id{state}$ recursively.
\marginpar{claim}
	So for every finite SHS that has been constructed from the initial state by means of event occurrences,
	the function $\id{state}$ is defined.
	This ensures that every event in $\Events$ has precisely one state, which means that states are defined predictably in a DIS.

@@

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} means that multiple primitive actions are executed as one.
	So let us first define the notion of atomic action.

>>>>

	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
	A {\em historic structure} is a tuple $\la \Events, \id{prec}\ra$ in which
\begin{eqnarray}
	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
	So for every event $e\in\Events$ we can say that it has happened (at some time in the past).
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\Events$.
	For this reason, requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
	meaning that $\id{prec}^*$ is reflexive, transitive, and antisymmetric.
	For this reason equation~\ref{req:precPlus asy} requires the transitive closure $\id{prec}^+$ to be antisymmetric.
	In the sequel, we will use predicate $\id{HS}\la \Events, \id{prec}\ra$
	to denote that $\la \Events, \id{prec}\ra$ is a historic structure
	and we will use variables $h$ and $h'$ to denote historic structures.

@@

	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
	Note that any nonempty linear historic structure has one current event.
	Other historic structures may have more.
	In a linear historic structure, every event except the first one has precisely one preceding event.
	In general, however, events may have multiple preceding events and multiple succeeding events.

@@

\begin{lemma}
\label{lemma:happens preserves historic structure}
	If $h=\la \Events, \id{prec}\ra$ is a historic structure and $P\subseteq\Events$ and $e\notin\Events$
	then $\id{happens}_{P,e}(h)$ is a historic structure.
\end{lemma}
	Lemma~\ref{lemma:happens preserves historic structure} says that $\id{happens}_{P,e}$ preserves the historic structure.
	This lemma has been proven in proof assistant Isabelle/HOL.
	To understand why $P\subseteq\Events$ is necessary,
	let $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
	However, this violates requirement~\ref{req:precInE}.
	So it is necessary that every $e_1\in P$ is an element of $\Events$.
	Similarly, $e\notin\Events$ is a necessary condition,
	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).

	Together with an empty historic structure, $\la \emptyset, \emptyset\ra$,
	we can define a historic structure recursively by means of $\id{happens}$:
\begin{eqnarray}
	&&\id{HS}\la \emptyset, \emptyset\ra\label{def:zeroHS}\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{HS}\la \Events, \id{prec}\ra\ \Rightarrow\ \id{HS}(\id{happens}_{P,e}\la \Events, \id{prec}\ra)
\end{array}
\label{req:recursive def happens}
\end{eqnarray}
	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a historic structure.
	Equation~\ref{req:recursive def happens} specifies recursively that $\id{happens}_{P,e}(h)$ is a historic structure iff
	$h$ is a historic structure and if $P$ and $e$ satisfy the specified conditions.
	This definition allows us to prove properties of finite historic structures by induction.
	In practical situations, every historic structure is finite.

@@

	Having defined historic structures and the way they grow, let us proceed to determine requirements for states.

@@

	Let $\mathbb S$ represent the set of all conceivable states.
	A {\em stateful historic structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
	$\la \Events, \id{prec}\ra$ is a historic structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
	that maps every event in $\Events$ to a state.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful historic structure.
	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
	Or shorter: $s$ is the state after event $e$.
	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
	(and therefore a set of pairs).
	A {\em current state} is the state after a current event
	and a {\em past state} is the state after a past event.

	In a stateful historic structure, the occurrence of an event $e$ defines a new historic structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
\end{array}	
\label{eqn:def occurs}
\end{equation}
	In fact, this definition expands the definition of historic structures (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves historic structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful historic structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	To describe the growth of a stateful historic structure we can define stateful historic structures recursively:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	Information systems can register changes (as opposed to states) to avoid duplicating most of a previous state into a new state.
	Such changes are called {\em transactions}.
	For the purpose of computing a new state ($s_2$) from a given state ($s_1$),
	we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	In a stateful historic structure we can represent transactions by means of a function $\delta$,
	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
\begin{equation}
	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
\label{def:delta}
\end{equation}
	The transaction caused by event $e$ can be described by:
\begin{equation}
	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
		\id{state}(e)\ =\ s&\wedge\\
		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
	\end{array}
\label{req:transaction}
\end{equation}
	We can define transactions recursively, in which we designate one specific state,
	$\mathcal I\in\mathbb S$,
	as initial state.
\begin{equation}
\begin{array}{l}
	e\notin\Events\ \wedge\ P\subseteq\Events\ \Rightarrow\\
	\id{trans}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
				&=&\mathcal I\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}
		\end{array}
\end{array}
\label{req:recursive transaction}
\end{equation}

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} means that a state change caused by a sequence of events can be reproduced as a single event.

	So let us first discuss event sequences.
	Let us define an event sequence as a totally ordered historic structure.
	Note that a historic structure that is not totally ordered may contain multiple historic substructures
	that are totally ordered.
	Typically, event sequences occur multiply in a historic structure.

@@
